---
title: "Accompanying codes for the project"
author: "Qichen He, Shin Ah Oh, Yue Shen, Mo Zhou"
date: "5/17/2020"
output:
  pdf_document: default
  html_document: default
---

## Preparing the data

```{r, warning = FALSE, message = FALSE}

## Load necessary packages and functions

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(rstan)
library(zoo)
library(loo)
library(bayesplot)
source("Helping_functions.R")

```

```{r}

## Load data from local file location

load("C:/Users/fairy/Documents/Bayesian Statistics/Project - Final Output/filnal code and result/filnal code and result/30263-0001-Data.rda")
data_full <- da30263.0001 



## Choose school code 105; same code applies to school code 104

data = get_school_data(data_full, 105) 



## Get the student id with 0 or 1 missing data

student_all_grade = data %>% group_by(case_identifier) %>% 
    summarise(count = sum(is.na(score))) %>% filter(count < 1)

```

We interpolate `NA`s using linear approximation.



## Generating the data exploration figures in the paper 

```{r}

## Relationship between math achievement and rates of change

data.figure1 = data %>% 
    filter(case_identifier %in% student_all_grade$case_identifier) %>%
    mutate(grade = as.factor(grade)) %>%
    mutate(initial = as.factor(ifelse(score < 55, 1, 
                            ifelse(score < 60, 2, 
                                   ifelse(score < 65, 3, 4))))) %>%
    group_by(initial, grade) %>%
    summarise(score = mean(score))

interaction.plot(data.figure1$grade, 
                 data.figure1$initial,
                 data.figure1$score, 
                 legend = F, xlab = "Grade",ylab = "Math Ach.")

rm(data.figure1)

```

```{r}

## Initial math achievement and rates of change with gender 

data.figure2 = data %>% 
    filter(case_identifier %in% student_all_grade$case_identifier) %>%
    mutate(grade = as.factor(grade),
           case_identifier = as.factor(case_identifier)) %>%
    mutate(initial = as.factor(ifelse(score < 60, 1, 2))) %>%
    group_by(initial, grade, Gender_coded) %>%
    summarise(score = mean(score)) 

ggplot(data.figure2[data.figure2$Gender_coded == 0,]) + 
    geom_line(aes(grade, score,group = initial), linetype = "dotdash") +
    geom_line(data = data.figure2[data.figure2$Gender_coded == 1,], 
              mapping = aes(grade, score,group = initial)) +
    theme_light()
 rm(data.figure2)
 
```



## We further tidy the data to be used for modeling

```{r}

data_model = data %>% 
  filter(case_identifier %in% student_all_grade$case_identifier) %>%
  group_by(case_identifier) %>% mutate(score = na.locf(score))

```



## Model building - Model 1

```{r}

## Build model

dat1 = get_dat1_list(data_model)

  # Load Stan file from local file location

fileName <- "C:/Users/fairy/Documents/Bayesian Statistics/Project - Final Output/model4.stan"
pars_1 = c("beta_0","beta_1","beta_01","beta_11","sigma_u0","sigma_u1","sigma")

  # Run model

resStan1 = train_two_level(fileName = fileName, data = dat1, pars = pars_1, iter = 10000, warmup = 3000)

  # Traceplot to check convergence

traceplot(resStan1, pars = pars_1,ncol = 2,inc_warmup = FALSE)

  # Print results

print(resStan1, pars = pars_1)



## Model checking

y_rep <- as.matrix(resStan1, pars = "y_rep")
dim(y_rep)

ppc_dens_overlay(dat1$Y_ti, y_rep[200:500, ]) + 
  theme_light() + 
  theme(legend.position="top")

p_value = mean(apply(y_rep,1,Test)>Test(dat1$Y_ti))
ppc_stat(dat1$Y_ti, yrep = y_rep, stat = Test)+
  theme_light() + 
  theme(legend.position="top")
p_value

```

Here, we compared the density of y with densities of y from 200 posterior draws.



## Model building - Model 2

```{r}

dat2 = get_dat2_list(data_model)

  # Load Stan file from local file location

fileName2 <- "C:/Users/fairy/Documents/Bayesian Statistics/Project - Final Output/model5.stan"
pars_2 = c("beta_0","beta_1","beta_01","beta_11","sigma_u0","sigma_u1","sigma","b")

  # Run model

resStan2 = train_two_level(fileName = fileName2, data = dat2, pars = pars_2,iter = 10000, warmup = 3000)

  # Traceplot to check convergence

traceplot(resStan2, pars = pars_2,ncol = 2,inc_warmup = FALSE)

  # Print results

print(resStan2, pars = pars_2)



## Model checking

y_rep <- as.matrix(resStan2, pars = "y_rep")
dim(y_rep)
ppc_dens_overlay(dat2$Y_ti, y_rep[1:200, ])

p_value = mean(apply(y_rep,1,Test)>Test(dat2$Y_ti))
ppc_stat(dat2$Y_ti, yrep = y_rep, stat = Test)

```



## Model building - Model 3

```{r}

dat3 = get_dat2_list(data_model)

  # Load Stan file from local file location

fileName <- "C:/Users/fairy/Documents/Bayesian Statistics/Project - Final Output/model6.stan"
pars_3 = c("beta_0","beta_1","beta_01","beta_11","sigma_u0","sigma_u1","sigma","b1","b2")

  # Run model

resStan3 = train_two_level(fileName = fileName, data = dat3, pars = pars_3,iter = 10000, warmup = 3000)

  # Traceplot to check convergence

traceplot(resStan3, pars = pars_3,ncol = 2,inc_warmup = FALSE)

  # Print results

print(resStan3, pars = pars_3)



## Model checking

y_rep <- as.matrix(resStan3, pars = "y_rep")
dim(y_rep)
ppc_dens_overlay(dat3$Y_ti, y_rep[1:200, ])

p_value = mean(apply(y_rep,1,Test)>Test(dat3$Y_ti))
ppc_stat(dat3$Y_ti, yrep = y_rep, stat = Test)
p_value

```



## Additional steps for school 105, Model 3 (this section does not apply to school 104)

```{r}

dat3 = get_dat2_list(data_model)

  # Load Stan file from local file location

fileName <- "C:/Users/fairy/Documents/Bayesian Statistics/Project - Final Output/model7.stan"
pars_3 = c("beta_0","beta_1","beta_01","beta_11","sigma_u0","sigma_u1","sigma","b1","b2")

  # Run model

resStan4 = train_two_level(fileName = fileName, data = dat3, pars = pars_3,iter = 10000, warmup = 3000)

  # Traceplot to check convergence

traceplot(resStan4, pars = pars_3,ncol = 2,inc_warmup = FALSE)

  # Print results

print(resStan4, pars = pars_3)

```



## Comparing Models 1, 2, and 3

```{r, warning = FALSE}

fitlist <- list(
  resStan1 = resStan1,
  resStan2 = resStan2,
  resStan3 = resStan3
)

fitlist  %>% 
  lapply(extract_log_lik) %>%
  lapply(loo) ->
  loolist
loolist
loolist[1:3] %>% loo_compare(x = .)

```


